---
author: "Jianhua Huang"
date: "July 23, 2016"
title: "Streamline Routine Modeling Work in R: streamlineR"
output:
  md_document:
    toc: true
    variant: markdown_github
---

# Streamline Routine Modeling Work in R: streamlineR
**For an interactive presentation version, please check [here](https://jianhua.shinyapps.io/streamlineR_shinyapp/)**


```{r,set.options, include=FALSE}
# dir <- 'F:/Projects/Rpackage/streamlineR'
dir <- 'C:/Users/Jianhua/Dropbox/work_doc/Rpackage/streamlineR'
knitr::opts_chunk$set(echo = TRUE, root.dir = dir, warning=FALSE, message=FALSE)
```


This package is designed to streamline the routine modeling work, especially for scoring. It provides some handy functions to bin numerical variables, replace numerical variables with Weight of Evidence (WOE), ranking variables by Information Values (IV), plotting the successful/failure rates, check model performance based on AUC, and so on.This package also provides the useful function to convert the model output (e.g., coefficients) to graph/tables that are easier to understand for non-technical audience.

The following example illustrates how to use the `streamlineR` package to prepare data, build models, and visualize results. 

## Packages Setup

#### Install Dependent Packages 
```{r, eval=FALSE}
# If the default mirror is blocked, choose another mirror to install R packages, 
# chooseCRANmirror()
sapply(c('dplyr', 'car', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart', 
  'scales', 'shiny', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS', 'pROC', 
  'manipulate'), 
  install.packages)
```

#### Load Dependent Packages 
```{r}
# Load pacakges
sapply(c('dplyr', 'car', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart', 
  'scales', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS', 'pROC', 
  'manipulate'), 
  require, character.only = TRUE)
```

#### Install `streamlineR` from Github 
```{r}
# Ff the install_github does not work, you can download the package from github,
# and then install the package locally: 
# https://api.github.com/repos/JianhuaHuang/streamlineR/zipball/master 
devtools::install_github('JianhuaHuang/streamlineR')
library(streamlineR)
```


## Data Preparation

#### Load Data 
```{r}
dt <- survival::pbc %>%
  transmute(age = round(age), gender = sex, platelet, stage = as.character(stage), 
    time, status = as.numeric(status %in% c(1, 2))) %>%
  filter(!is.na(stage))

set.seed(1111)  # reproducible
dt <- dt[sample(nrow(dt), 10000, replace = T), ]
row.names(dt) <- NULL
dim(dt)
str(dt)
head(dt)
```

#### Split Data into Training and Test Datasets 
```{r}
set.seed(1111)
ind.train <- createDataPartition(dt$status, p = .7, list = FALSE)
dt.train <- dt[ind.train, ]
dt.test <- dt[-ind.train, ]
row.names(dt.train) <- NULL
row.names(dt.test) <- NULL
dim(dt.train)
dim(dt.test)

# preserve the original values
dt.train.bk <- dt.train
dt.test.bk <- dt.test
```



## Bin Training Data Based on Regression Coefficients: `bin.knn`

```{r}
bin.knn(status ~ platelet, data = dt.train, n.group = 5, min.bucket = 0.05)
```


```{r,eval=FALSE}
manipulate(bin.knn(status ~ platelet, data = dt.train, n.group, min.bucket),
  n.group = slider(1, 10, step = 1, initial = 5, label = 'Number of Groups'),
  min.bucket = slider(0.01, .2, step = 0.01, initial = 0.05,
    label = 'Minimum Population'))
```

## Bin Training Data Based on rpart: `bin.rpart` 

#### Decision Tree Algorithm (Recursive Partitioning): `rpart`
```{r}
rpart(formula = status ~ age, data = dt.train, 
  control = rpart.control(minbucket = .01 * nrow(dt.train)))
```

#### Binning for Logistic Model
```{r}
lg.bin.age <- bin.rpart(formula = status ~ age, data = dt.train, 
  rcontrol = rpart.control(minbucket = .01 * nrow(dt.train)))

str(lg.bin.age)

lg.bin.platelet <- bin.rpart(formula = status ~ platelet, data = dt.train, 
  rcontrol = rpart.control(minbucket = .01 * nrow(dt.train)))
```


#### Binning for Survival Model
```{r}
surv.bin.age <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)))  ## cp = 0.01
```


#### Search for Appropriate Number of Cut Points
```{r}
surv.bin.age2 <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)), n.group = 3:7)
```


#### Replace Numerical Varialbes with Bins
```{r}
# We don't need the time column anmore, delete it in both dt.train and dt.test
dt.train <- dplyr::select(dt.train, -time)  
dt.test <- dplyr::select(dt.test, -time)
head(dt.train)
dt.train$age <- lg.bin.age$bins
dt.train$platelet <- lg.bin.platelet$bins
head(dt.train)
```



## Level Statistics (Frequence, Rate, WOE, and IV): `level.stat`

```{r}
col.x <- c('age', 'gender', 'platelet', 'stage')
stat.train <- level.stat(dt.train, x = col.x, y = 'status')
head(stat.train)
```


## Visualize Level Statistics: `ggstat`

#### Plot with Default Arguments
```{r}
ggstat(data = stat.train, var = 'Variable.IV', x = 'Group', y = 'Rate.1',
  y.label = 'Perc.1', y.min.0 = FALSE, y.title = NULL, bar.width = 'Rate.group',
  bar.width.label = 'Perc.group', n.col = NULL)
```

#### Constant Bar Width
```{r}
ggstat(stat.train, bar.width = NULL)
```

#### Plot WOE
```{r,fig.height=3}
ggstat(stat.train, y = 'WOE', y.label = 'WOE.round', bar.width = NULL, 
  bar.width.label = NULL, n.col = 4)
```


## Replace Bins with WOE: `replace.woe`

```{r}
replace.woe(data = dt.train, level.stat.output = stat.train, replace = FALSE) %>%
  head

dt.train <- replace.woe(data = dt.train, level.stat.output = stat.train, 
  replace = TRUE)
head(dt.train)
```


## Correlation between Independent Variables: `corrplot.beautify`

```{r}
cor.mat <- cor(dt.train[, col.x])
corrplot(cor.mat)
corrplot.beautify(cor.mat)
```


## Logistic Model

#### Full Model
```{r}
lg <- glm(status ~ ., dt.train, family=binomial(link='logit'))
summary(lg)
```

#### Stepwise Variable Selection
```{r}
lg.aic <- stepAIC(lg, k =  qchisq(0.05, 1, lower.tail=F))   # p to enter: 0.05
summary(lg.aic)

data.frame(vif(lg.aic))
```


## Prepare Test Data: `bin.custom & replace.woe`

#### Bin Test Data: `bin.custom`
```{r}
head(dt.test)
dt.test$age <-  bin.custom(dt.test$age, cut.p = lg.bin.age$cut.points)
dt.test$platelet <- bin.custom(dt.test$platelet, cut.p = lg.bin.platelet$cut.points)
head(dt.test)
```


#### Replace Binned Test Data with WOE: `replace.woe`
```{r}
dt.test <- replace.woe(dt.test, level.stat.output = stat.train, replace = TRUE)
head(dt.test)
```


## Model Performance: `perf.auc & perf.decile`

#### Check Performance Based on AUC: `perf.auc`
```{r}
perf.auc(model = lg.aic, dt.train, dt.test)
```


#### Check Performance Based on Decile Rate: `perf.decile`
```{r}
pred.test <- predict(lg.aic, newdata = dt.test, type = 'response')
perf.decile(actual = dt.test$status, pred = pred.test, add.legend = TRUE)
```



## Convert Coefficients to Rate: `coef2rate`

```{r}
summary(lg.aic)
pred.stat <- coef2rate(data = dt.test, model = lg.aic, 
  level.stat.output = stat.train, force.change = TRUE)
head(pred.stat)
# Compare The Actual & Prediction Rates
pred.stat[,c('Rate.1', 'Pred.Rate.1')]

ggstat(pred.stat, y = 'Pred.Rate.1')
```


## Reference:
* streamlineR package information: https://github.com/JianhuaHuang/streamlineR 
* Submit bug: http://github.com/JianhuaHuang/streamlineR/issues
