---
title: "Untitled"
author: "Jianhua Huang"
date: "July 23, 2016"
output:
  md_document:
    variant: markdown_github
---

This package is designed to streamline the routine modeling work, especially for scoring. It provides some handy functions to bin numerical variables, replace numerical variables with Weight of Evidence (WOE), ranking varialbes by Information Values (IV), plotting the successful/failure rates, check model performance based on AUC, and so on.This package also provides the useful function to convert the model output (e.g., coefficients) to graph/tables that are easier to understand for non-technical audience.

R Setup
```{r setup, warning=FALSE, message=FALSE, include=FALSE}
rm(list = ls())
# If you don't have these packages, install them. 
# once the packages are installed, you don't need to run the code again
# choose mirror to download R packages, in case the default one is blocked
# chooseCRANmirror()  
# sapply(c('dplyr', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart',
#   'scales', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS', 'pROC'),
#   install.packages)

# Load packages
sapply(c('dplyr', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart', 
  'scales', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS', 'pROC'), 
  require, character.only = TRUE)

# install the streamit package via github
# install_github('JianhuaHuang/streamit')  # remove the "#" at to run this line
library(streamit)

# dir <- 'F:/Projects/Rpackage/streamit'
dir <- 'C:/Users/Jianhua/Dropbox/work_doc/Rpackage/streamit'

opts_chunk$set(echo = TRUE, root.dir = dir, warning=FALSE, message=FALSE)
```


## Data Preparation
```{r}
dt <- survival::pbc %>%
  transmute(age = round(age), gender = sex, platelet, stage = as.character(stage), 
    time, status = as.numeric(status %in% c(1, 2))) %>%
  filter(!is.na(stage))

set.seed(1111)
dt <- dt[sample(nrow(dt), 10000, replace = T), ]
dim(dt)
str(dt)
head(dt)
```


## Split Data into Training and Test datasets
```{r}
# set.seed(123456)
set.seed(1111)
ind.train <- createDataPartition(dt$status, p = .7, list = FALSE)
dt.train <- dt[ind.train, ]
dt.test <- dt[-ind.train, ]
row.names(dt.train) <- 1:nrow(dt.train)
row.names(dt.test) <- 1:nrow(dt.test)
dim(dt.train)
dim(dt.test)
rm(dt)
```


## Binning Based on rpart: `bin.rpart`
```{r}
rpart(formula = status ~ age, data = dt.train, 
  control = rpart.control(minbucket = .05 * nrow(dt.train)))
```

**Binning for Logistic Model**
```{r}
lg.bin.age <- bin.rpart(formula = status ~ age, data = dt.train, 
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)))

str(lg.bin.age)

lg.bin.platelet <- bin.rpart(formula = status ~ platelet, data = dt.train, 
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)))
```

**Binning for Survival Model**
```{r}
surv.bin.age <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)))  ## cp = 0.01

surv.bin.age <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(cp  = .001, minbucket = .05 * nrow(dt.train)))  

surv.bin.age2 <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)), n.group = 3:7)
```

**Replace numerical Varialbes with Bins**
```{r}
dt.train <- dplyr::select(dt.train, -time)
dt.test <- dplyr::select(dt.test, -time)
head(dt.train)
dt.train$age <- lg.bin.age$bins
dt.train$platelet <- lg.bin.platelet$bins
head(dt.train)
```


## Level Statistics (Frequence, Rate, WOE, and IV): `level.stat`
```{r}
col.x <- c('age', 'gender', 'platelet', 'stage')
stat.train <- level.stat(dt.train, x = col.x, y = 'status')
head(stat.train)
```


## Visualizing Level Statistics: `ggstat`
```{r}
ggstat(data = stat.train, var = 'Variable.IV', x = 'Group', y = 'Rate.1',
  y.label = 'Perc.1', y.min.0 = FALSE, y.title = NULL, bar.width = 'Rate.group',
  bar.width.label = 'Perc.group', n.col = NULL)

ggstat(stat.train)
```


**Constant Bar Width**
```{r}
ggstat(stat.train, bar.width = NULL)
```


**Plot WOE**
```{r,fig.height=3}
stat.train$WOE.round <- round(stat.train$WOE, 2)
ggstat(stat.train, y = 'WOE', y.label = 'WOE.round', bar.width = NULL, 
  bar.width.label = NULL, n.col = 4)
```


## Replace Bins with WOE: `replace.woe`
```{r}
replace.woe(data = dt.train, level.stat.output = stat.train) %>% head

dt.train.org <- dt.train
dt.train <- replace.woe(data = dt.train, level.stat.output = stat.train, 
  replace = TRUE)
head(dt.train)
```


## Correlation between Independent Variables: `corrplot.beautify`
```{r}
cor.mat <- cor(dt.train[, col.x])
corrplot.beautify(cor.mat)
```


## Logistic Model
```{r}
lg <- glm(status ~ ., dt.train, family=binomial(link='logit'))
summary(lg)

lg.aic <- stepAIC(lg, k =  qchisq(0.05, 1, lower.tail=F))   # p to enter: 0.05
summary(lg.aic)
```


## Preparing Test Data: `bin.custom & replace.woe`
**Bin Test Data**: `bin.custom`
```{r}
head(dt.test)
dt.test$age <-  bin.custom(dt.test$age, cut.p = lg.bin.age$cut.points)
dt.test$platelet <- bin.custom(dt.test$platelet, cut.p = lg.bin.platelet$cut.points)
head(dt.test)
```


**Replace Binned Test Data with WOE**: `replace.woe`
```{r}
dt.test <- replace.woe(dt.test, level.stat.output = stat.train, replace = TRUE)
head(dt.test)
```


## Model Performance: `perf.auc & perf.decile`
**Check Performance Based on AUC**: `perf.auc`
```{r}
perf.auc(model = lg.aic, dt.train, dt.test)
```

**Check Performance Based on Decile Rate**: `perf.decile`
```{r}
pred.test <- predict(lg.aic, newdata = dt.test, type = 'response')
perf.decile(actual = dt.test$status, pred = pred.test, add.legend = TRUE)
```


## Convert Coefficients to Rate: `coef2rate`
```{r}
pred.stat <- coef2rate(data = dt.test, model = lg.aic, 
  level.stat.output = stat.train, force.change = TRUE)
head(pred.stat)
pred.stat[,c('Rate.1', 'Pred.Rate.1')]

ggstat(pred.stat, y = 'Pred.Rate.1')
```

