---
title: 'Streamline Routine Modeling Work in R: streamlineR'
author: "Jianhua Huang"
date: "July 27, 2016"
output: 
  slidy_presentation:
    fig_width: 7
    fig_height: 5
    duration: 60
    incremental: true
runtime: shiny
---

```{r knitr.setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


```{r sourcecode, include = FALSE}
#   Build and Reload Package:  'Ctrl + Shift + B'
#   Check Package:             'Ctrl + Shift + E'
#   Test Package:              'Ctrl + Shift + T'

############################# little functions #################################
## add.row
add.row <- function(df, new.row) {
  # This function is used to add a new row to the end of a data.frame.
  # Each element of the new row has the same class as the column it belongs to
  # However, factor will be convert to character firstly
  # Args:
  #   df: The data.frame to be added to
  #   new.row: a vector of the new values added to the end of df.
  #            The length of the new.row should be equal to number of df columns
  # Return:
  #   df with new row added to the end. The classes of each column is preserved
  f <- sapply(df, is.factor)
  df[f] <- lapply(df[f], as.character)
  
  df.class <- lapply(df, class)
  new.row <- as.list(new.row)
  if(length(df.class) != length(new.row)) {
    stop('Length does not match number of columns')
  }
  
  new.row.list <- lapply(1:length(new.row), function(x) {
    new.class <- as(new.row[[x]], df.class[[x]])
  })
  
  df[nrow(df) + 1, ] <- new.row.list
  
  return(df)
}
# df <- data.frame(x = c('A', 'B', 'C'), y = 1:3, z = 4:6, stringsAsFactors = F)
# df <- data.frame(x = c('A', 'B', 'C'), y = 1:3, z = 4:6)
# new.row <- c('Total', colSums(df[, -1]))
# add.row(df = df, new.row = new.row)

## paste.mat
# paste elements to a matrix, and return a matrix with the same dims
paste.mat <- function(x, ...) {
  # This function is used to paste matrix/data.frame, and return a matrix with
  # the same dim as x.
  # Arg:
  #   x: the first matrix/data.frame, which is used to determine the dimension
  #   ...: any other elements that are being pasted to the x, it can be a
  #        single value, a vector, and a matrix. (data.frame must be converted
  #        to matrix before being passed into the function)
  # Return:
  #   a matrix with the same dimension as x, column name reserved.
  if(is.data.frame(x)) x <- as.matrix(x)
  nrow.x <- nrow(x)
  value <- paste0(x, ...)
  out.mat <- matrix(value, nrow = nrow.x)
  colnames(out.mat) <- colnames(x)
  return(out.mat)
}
# df <- data.frame(a=1:3, b = 4:6, c = 7:9)
# paste.mat(df, '(', as.matrix(df), ')')
# paste.mat(df, '_', 11:19)

# ifelse prevent data type change
ifelse.safe <- function(cond, yes, no) {
  # The ifelse function may change the returned data format (e.g., date -> num)
  # this simple function can prevent the change
  # reference to:
  # http://stackoverflow.com/questions/6668963/how-to-prevent-ifelse-from-turning-
  # date-objects-into-numeric-objects
  structure(ifelse(cond, yes, no), class = class(no))
}

############################## Binning and WOE #################################
bin.custom <- function(x, cut.p, names = NULL,
  name.style = c('report', 'scientific'), ...) {
  # This function is used to bin the data with given cut points
  # Args:
  #    x: a single column of data to be cut
  #    cut.p: the given cut points.
  #    names: the names given to the bins after cutting. If no names is
  #               given, the default name.style is used
  #    name.style: the naming style used, can be either
  #                'scientific' (e.g., (1, 2]) or 'report' (e.g., 1 < · <= 2)
  # Return:
  #    A vector of binned values
  
  x.num <- as.numeric(x)
  cut.p.num <- as.numeric(cut.p)
  
  # add ends to the cut.p
  cut.p_ends <- c(-Inf, cut.p.num, max(x.num, na.rm = T))
  x.bins <- cut(x.num, cut.p_ends, include.lowest = T)
  
  # change the levels from -Inf to the actual minimum value
  if(is.null(names)) {
    
    name.style = name.style[1]
    
    if(name.style == 'scientific') {
      levels(x.bins) <- gsub(paste0('[-Inf,', cut.p[1], ']'),
        paste0('[', min(x, na.rm = T), ',', cut.p[1], ']'),
        levels(x.bins), fixed = T)
    }
    
    if(name.style == 'report') {
      bin.names <- c(paste('<=', cut.p[1]), rep(NA, length(cut.p) - 1),
        paste('>', cut.p[length(cut.p)]))
      
      if(length(cut.p) > 1) {
        for (i in 2:length(cut.p)) {
          bin.names[i] <- paste(cut.p[i - 1], '< \u00B7 <=', cut.p[i])
        }}  # \u00B7 is the unicode for mid-dot, \u2022 is for bullet point
      levels(x.bins) <- bin.names
    }
  }
  # if there is only one unique value in the level, change the x.bins to
  # that value
  x.bin.unique <- data.frame(x, x.bins) %>%
    filter(!is.na(x)) %>%
    group_by(x.bins) %>%
    summarise(unique.x = length(unique(x)), min.x = min(x))
  
  levels(x.bins) <- ifelse(x.bin.unique$unique.x == 1, x.bin.unique$min.x,
    levels(x.bins))
  
  if(!is.null(names)) levels(x.bins) <- names # coerce to change levels
  
  # preserve the levels order
  lvs <- levels(x.bins)
  x.bins <- as.character(x.bins)
  x.bins[is.na(x.bins)] <- 'Missing'
  x.bins <- factor(x.bins, levels = c(lvs, 'Missing'))
  x.bins <- droplevels(x.bins)
  return(x.bins)
}
# df <- data.frame(x = 1:100, y = sample(c(1, 0), 100, replace = T))
# df <- data.frame(x = 1:100, y = rep(c(1, 0), each = 50))
# bin.custom(x = c(1:50, NA, NA, 51:100), cut.p = c(1, 2, 99), name.style = 'report')
# rs <- bin.custom(dt$Grad_Year_Month, as.yearmon(c('Aug 2012', 'Aug 2013')))

## bin.knn
bin.knn <- function(formula, data, n.group, min.bucket) {
  # Visualize the binning for survival/logistic model based on model coefficients
  # Can be combined with the manipulate function to check the binning of 
  # coefficients interactively
  # Args:
  #    formula: the model used for binning (can be glm or coxph)
  #    data: the data frame used for binning
  #    n.group: number of binned groups
  #    min.bucket: the minimum ratio of population in each group (can be 
  #             a value between 0 to 1)
  # Return:
  #    Shows a ggplot with the regression coefficients and the binned groups
  
  x <- as.character(formula[3])
  y <- as.character(formula[2])
  
  # This function is used to explore the clustering of bins based on KNN method
  cut.q <- unique(quantile(data[, x], seq(0, 1, length.out = 1 / min.bucket),
    na.rm = T))
  data[, x] <- cut(data[, x], cut.q, include.lowest = TRUE)
  data[, x] <- droplevels(data[, x])
  data[, x] <- addNA(data[, x], ifany = T)
  
  # If y is a survival object, coxph is used to estimate the coefficients
  # Otherwise (y is a binary vector), the logistic model is used
  if(grepl('^Surv', y)) {
    mod <- coxph(formula, data = data)
  } else {
    mod <- glm(formula, data = data, family=binomial(link='logit'))
  }
  
  # construct the coef and population dataframe
  x.pop <- data.frame(Variable = x, data.frame(table(data[, x])))
  colnames(x.pop) <- c('Variable', 'Bin', 'Freq')
  
  coef <- data.frame(Variable = x,
    Bin = gsub(x, '', names(mod$coefficients)),
    Coef = mod$coefficients)
  
  coef.pop <- merge(x.pop, coef, all.x = T)
  coef.pop$Coef[1] <- 0  # the first value is the reference, its coef is zero
  
  # knn groups based on the order and coefficients
  x.knn <- data.frame(Ord = 1:nrow(coef.pop), Coef = coef.pop$Coef)
  
  coef.pop$KNN_Group <- as.factor(
    kmeans(scale(x.knn), n.group, nstart = 20)$cluster)
  
  ggplot(coef.pop, aes(x = Bin, y = Coef, fill = KNN_Group,
    color = KNN_Group, width = 3 * Freq / nrow(data))) +
    geom_bar(stat = 'identity') +
    labs(x = 'Bins', y = 'Model Coefficients',
      title = paste(x, ':', nrow(coef.pop), 'coefficients clustered into',
        n.group, 'groups')) +
    theme_bw() +
    scale_fill_discrete(name = 'KNN Group') +
    scale_color_discrete(name = 'KNN Group') +
    theme(axis.text.x = element_text(angle=30, hjust=1))
}
# manipulate(bin.knn(status ~ platelet, data = dt.train, n.group, min.bucket),
#   n.group = slider(1, 10, step = 1, initial = 5, label = 'Number of groups'),
#   min.bucket = slider(0.01, 1, step = 0.01, initial = 0.05,
#     label = 'Minimum Population Size (%)'))

## rpart.bin
bin.rpart <- function(formula, data, rcontrol = rpart.control(), n.group = NULL, ...) {
  # This function is used to bin the numerical variable for survival model
  # Arg:
  #    formula: the formula used for rpart
  #    data: the dataset used for rpart
  #    n.group: the acceptable number of groups (NA group not counted for)
  #    rcontrol: the control used for rpart
  
  # The NA values are removed by the rpart function automatically
  row.names(data) <- 1:nrow(data)
  
  vars <- all.vars(formula)
  x.num <- vars[length(vars)]
  
  # if the minbucket is the default value 7, then update it to 1% of the data
  # if(rcontrol$minbucket == 7) {
  #   rcontrol$minbucket <- .01 * nrow(data)
  # }
  
  rp.tree <- rpart(formula, data, control = rcontrol, ...)
  # rp.tree <- rpart(formula, data, control = rcontrol)
  
  ## if n.group is NULL, and no group is found, return 'No Bin'
  if(is.null(n.group) & length(unique(rp.tree$where)) == 1) {
    cat(c(x.num, ': No Bin \n'))
    return('No Bin')
  }
  
  ## if n.group is not NULL, change cp to find the possible bins within n.group
  while((!is.null(n.group)) & (!length(unique(rp.tree$where)) %in% n.group)) {
    multipler <- ifelse(length(unique(rp.tree$where)) > median(n.group), 1.1, .9)
    rcontrol$cp <- rcontrol$cp * multipler
    rp.tree <- rpart(formula, data, control = rcontrol, ...)
    # rp.tree <- rpart(formula, data, control = rcontrol)
  }
  
  tree.where <- data.frame(Where = rp.tree$where)
  
  tree.value <- data.frame(Value = data[, x.num], Where = 'Missing',
    stringsAsFactors = F)
  tree.value[row.names(tree.where), 'Where'] <- tree.where$Where
  
  tree.cut <- group_by(tree.value, Where) %>%
    summarise(Cut_Start = min(Value), Cut_End = max(Value)) %>%
    arrange(Cut_End)
  
  if(is.na(tree.cut$Cut_End[nrow(tree.cut)])) {
    cut.p <- tree.cut$Cut_End[1:(nrow(tree.cut) - 2)]
  } else {
    cut.p <- tree.cut$Cut_End[1:(nrow(tree.cut) - 1)]
  }
  
  cat(c(x.num, ':', cut.p, '\n'))
  
  bin.names <- c(paste('<=', cut.p[1]), rep(NA, length(cut.p) - 1),
    paste('>', cut.p[length(cut.p)]))
  
  if(length(cut.p) > 1) {
    for (i in 2:length(cut.p)) {
      bin.names[i] <- paste(cut.p[i - 1], '< \u00B7 <=', cut.p[i])
    }}  # \u00B7 is the unicode for mid-dot, \u2022 is for bullet point
  
  # check whether the Cut_Start and Cut_End are the same
  # if the same, the <, =, or > signs is not needed
  bin.names <- ifelse(tree.cut$Cut_Start[1:length(bin.names)] ==
      tree.cut$Cut_End[1:length(bin.names)], tree.cut$Cut_End, bin.names)
  
  tree.cut$Bin <- 'Missing'
  tree.cut$Bin[1:length(bin.names)] <- bin.names
  
  x.bins <- factor(tree.cut$Bin[match(tree.value$Where, tree.cut$Where)],
    levels = tree.cut$Bin)
  
  return(list(cut.points = cut.p, bins = x.bins))
}
# dt <- read.csv('C:/Projects/AlumniConvProg/data/Data_Associate.csv')
# bin.rpart(formula = Surv(Conversion_Time_Months, Conversion_Status) ~ WF_Count,
#   rcontrol = rpart.control(cp = 0.0001, minbucket = nrow(dt.conv) * .01),
#   data = dt.conv, n.group = 3:7)

## Rates by level
level.stat <- function(data, x = NULL, y, flag.0 = 0, flag.1 = 1) {
  # Used to calculate the rates, WOE, and IV for different variables
  # Args:
  #    data: the input data.frame
  #    x: the independent variables, can be a single variable or a vector
  #       including multiple variables
  #    y: the binary flag variable (e.g., Progression_Flag)
  #    flag.0: the value representing 0 in y
  #    flag.1: the value representing 1 in y
  # Return: a data.frame inclduing frequency, rates, WOE, and IV for each x level
  
  if(is.null(x)) x <- setdiff(colnames(data), y)
  
  rs.x <- lapply(x, function(xx) {
    # check whether Y is binary value
    if(length(unique(data[, y])) != 2) stop('Y is not binary value')
    
    dt <- data[, c(xx, y)]
    
    if(is.factor(dt[, y])) dt[, y] <- as.character(dt[, y])
    
    dt[, y] <- ifelse(dt[, y] == flag.0, 0, 1)
    
    dt.iv <- table(dt) %>%
      as.data.frame.matrix %>%
      transmute(Variable = xx,
        Group = row.names(.),
        Freq.0 = `0`,
        Freq.1 = `1`,
        Freq.group = Freq.0 + Freq.1,
        Rate.0 = Freq.0 / Freq.group,
        Rate.1 = Freq.1 / Freq.group,
        Rate.group = Freq.group / sum(Freq.group),
        Perc.0 = ifelse(Rate.0 > .01,
          paste0(round(Rate.0 * 100), '%'),
          paste0(round(Rate.0 * 100, 1), '%')),
        Perc.1 = ifelse(Rate.1 > .01,
          paste0(round(Rate.1 * 100), '%'),
          paste0(round(Rate.1 * 100, 1), '%')),
        Perc.group = ifelse(Rate.group > .01,
          paste0(round(Rate.group * 100), '%'),
          paste0(round(Rate.group * 100, 1), '%')),
        Distr.0 = Freq.0 / sum(Freq.0),
        Distr.1 = Freq.1 / sum(Freq.1),
        WOE = log(Distr.1 / Distr.0),
        WOE.round = round(WOE, 2), 
        IV = sum((Distr.1 - Distr.0) * WOE))
  })
  
  rs <- do.call(rbind, rs.x) %>%
    arrange(desc(IV)) %>%
    mutate(Variable.IV = paste0(Variable, ' (IV: ', round(IV, 3), ')'),
      Variable = factor(Variable, levels = unique(Variable)),
      Variable.IV = factor(Variable.IV, levels = unique(Variable.IV)),
      Group = factor(Group,
        levels = c(setdiff(unique(Group), 'Missing'), 'Missing')))
  return(rs)
}

## ggplot the level.stat output by variable and level
ggstat <- function(data, var = 'Variable.IV', x = 'Group', y = 'Rate.1',
  y.label = 'Perc.1', y.min.0 = FALSE, y.title = NULL, bar.width = 'Rate.group',
  bar.width.label = 'Perc.group', n.col = NULL) {
  # Plot the stat (statistics, e.g., Rate.1, Rate.0, and WOE) for each varaible
  # Args:
  #    data: the input dataset
  #    var: the varaibles plotted in different panel
  #    x: the variable used as x axis
  #    y: the varaible used as y axis
  #    y.label: the text used to label the y value
  #    y.min.0: whether to plot the bar from 0
  #    y.title: the title for y axis
  #    bar.width: the column used to represent bar width. If NULL, the bar width
  #               is set to 0.1
  #    bar.width.label: the column used to label the bar.width. If NULL, it is
  #                     not labelled.
  #    n.col: number of panel column
  # Return: ggplot
  
  data$var = data[, var]
  data$x = data[, x]
  data$y = data[, y]
  
  if(is.null(y.label)) {
    data$y.label = ''
  } else {
    data$y.label = data[, y.label]
  }
  
  if(is.null(bar.width)) {
    data$bar.width = 0
  } else {
    data$bar.width = data[, bar.width]
  }
  
  if(is.null(bar.width.label)) {
    data$bar.width.label = ''
  } else {
    data$bar.width.label = data[, bar.width.label]
  }
  
  if(is.null(n.col)) n.col <- ceiling(sqrt(length(unique(data$var))))
  
  y.range <- max(data$y) - ifelse(y.min.0 == TRUE, 0, min(data$y))
  y.min <- ifelse(y.min.0 == TRUE, 0, min(data$y) - y.range * .15)
  y.max <- (max(data$y)) + y.range * .1
  
  ggplot(data, aes(x = x, y = y)) +
    facet_wrap(~ var, scale = 'free', ncol = n.col) +
    geom_bar(aes(width = bar.width + .1), stat = 'identity',
      fill = 'cornflowerblue', color = 'cornflowerblue') +
    # geom_text(aes(y = y, label = y.label), size = 3, color = 'red',
    #   vjust = ifelse(data$y > 0, -.25, .85)) +
    geom_text(aes(y = y, label = y.label), size = 3, color = 'red',
      nudge_y = ifelse(data$y > 0, .06, -.06) * y.range) +
    geom_text(aes(y = y.min, label = bar.width.label), size = 3, vjust = 0) +
    labs(x = NULL, y = y.title) +
    scale_y_continuous(limits = c(y.min, y.max), oob = rescale_none) +
    theme_classic() +
    theme(axis.text.x = element_text(angle=25, hjust=1),
      axis.line.x = element_line(), axis.line.y = element_line(),
      axis.title = element_text(size=12,face="bold"),
      text = element_text(size = 10), strip.text = element_text(face = 'bold'),
      strip.background = element_blank())
}
# ggstat(lvs, y = 'Rate.1', y.label = 'Perc.1', bar.width = NULL)
# ggstat(lvs, y = 'WOE', y.label = NULL)
# ggstat(level.stat.prog, y.title = 'Progression Rate (%)')

## replace the categorical levels with their corresponding woe values
replace.woe <- function(data, level.stat.output, replace = FALSE) {
  # This function is used to replace the categorical variables with the
  # corresponding woe value claculated with level.stat function
  # Args:
  #    data: the input data.frame
  #    level.stat.output: the data.frame output from the level.stat function
  #    replace: FALSE/TRUE. If FALSE, the woe values will be added to the end.
  #             if TRUE, the original levels will be replaced by woe values
  # Return: a data.frame with woe
  
  dt.rp <- data[, intersect(colnames(data), level.stat.output$Variable)]
  dt.rp.woe <- lapply(colnames(dt.rp), function(x) {
    woe.x <- level.stat.output[level.stat.output$Variable == x, ]
    group.match <- match(dt.rp[, x], woe.x$Group)
    woe.x$WOE[group.match]
  })
  
  dt.rp.woe <- do.call(cbind, dt.rp.woe)
  
  if(replace == FALSE) {
    colnames(dt.rp.woe) <- paste0(colnames(dt.rp), '_woe')
    return(data.frame(data, dt.rp.woe))
  } else {
    data[, colnames(dt.rp)] <- dt.rp.woe
    return(data)
  }
}

## KM curve
km.curve <- function(data, time, status, x, plot = TRUE) {
  dt.km.obo <- lapply(x, function(xx) {
    sf.x <- survfit(as.formula(paste0('Surv(', time, ',', status, ') ~', xx)),
      data = data)
    df.x <- data.frame(time = sf.x$time, surv = sf.x$surv, variable = xx,
      group = rep(gsub(paste0(xx, '='), '', names(sf.x$strata)), sf.x$strata))
  })
  
  dt.km.obo <- do.call(rbind, dt.km.obo)
  
  out <- by(data = dt.km.obo, INDICES = dt.km.obo$variable, FUN = function(m) {
    m <- droplevels(m)
    m <- ggplot(m, aes(x = time, y = surv, color = group)) +
      geom_line() +
      scale_color_discrete(name = '') +
      labs(x = NULL, y = NULL) +
      facet_wrap(~variable, nrow = 4, scales = 'free') +
      theme_simple(plot.margin=unit(c(.5, 0, .5, .5), "cm"))
  })
  
  if(plot == TRUE) do.call(grid.arrange, out)
  return(dt.km.obo)
}
# sf <- km.curve(data = dt.conv, time = 'Conversion_Time_Months',
#   status = 'Conversion_Status', x = col.x[1:6], plot = TRUE)


## check model performance: perf.auc & perf.decile
perf.auc <- function(model, train, test) {
  # check the coxph or logistic (glm) model performance based on AUC
  # For coxph model, the time-dependent AUC and the iAUC (integrated AUC)
  # is calculated and plotted. For logistic (glm) model, the ROC curve is plotted
  # Args:
  #    model: the coxph or logistic model based on full dataset
  #    train: the training dataset
  #    test: the test dataset
  # Return: A plot of time-dependet AUC (for coxph), or ROC (for logistic)
  
  mod.class <- class(model)[1]  # glm model belongs to two classes, use the 1st
  if(!mod.class %in% c('coxph', 'glm')) {
    stop('the model class is not "coxph" or "glm"')
  }
  
  mod.train <- update(model, data = train)
  
  pred.train <- predict(mod.train,
    type = ifelse(mod.class == 'coxph', 'lp', 'response'))
  
  pred.test <- predict(mod.train, newdata = test,
    type = ifelse(mod.class == 'coxph', 'lp', 'response'))
  
  if(mod.class == 'coxph') {
    # use the time-dependent AUC for coxph model
    time <- all.vars(model$formula)[1]
    status <- all.vars(model$formula)[2]
    
    auc.train <- risksetAUC(Stime = train[, time], status = train[, status],
      marker = pred.train, tmax = max(train[, time]), plot = FALSE)
    
    auc.test <- risksetAUC(Stime = test[, time], status = test[, status],
      marker = pred.test, tmax = max(test[, time]), plot = FALSE)
    
    auc <- rbind(data.frame(auc.train, Data = 'Train'),
      data.frame(auc.test, Data = 'Test'))
    auc$Data <- factor(auc$Data, levels = unique(auc$Data))
    
    p <- ggplot(auc, aes(x = utimes, y = AUC, color = Data)) +
      geom_line(size = 1.5) +
      scale_size_continuous(guide = F) +
      scale_color_discrete(name = NULL,
        labels = c(paste0('Train (iAUC: ', round(auc.train$Cindex, 3), ')'),
          paste0('Test (iAUC: ', round(auc.test$Cindex, 3), ')'))) +
      theme_ws(legend.position = c(.2, .85)) +
      labs(x = 'Survival Time', y = 'AUC') +
      scale_y_continuous(limits = c(.5, 1))
  }
  
  if(mod.class == 'glm') {
    col.y <- all.vars(mod.train$formula)[1]
    
    roc.train <- roc(actual.train ~ pred.train,
      data = data.frame(actual.train = train[, col.y], pred.train))
    roc.train$auc
    
    roc.test <- roc(actual.test ~ pred.test,
      data = data.frame(actual.test = test[, col.y], pred.test))
    roc.test$auc
    
    ## ROC and AUC compare
    roc <- data.frame(
      FP = 1-roc.train$specificities,
      TP = roc.train$sensitivities,
      AUC = as.numeric(roc.train$auc),
      Data = 'Train') %>%
      rbind(data.frame(
        FP = 1-roc.test$specificities,
        TP = roc.test$sensitivities,
        AUC = as.numeric(roc.test$auc),
        Data = 'Test')) %>%
      mutate(Data = factor(Data, levels = unique(Data)))
    
    p <- ggplot(roc, aes(x = FP, y = TP, color = Data)) +
      geom_line() +
      scale_color_discrete(name = NULL, labels = c(
        paste0('Train (AUC: ', round(roc.train$auc, 3), ')'),
        paste0('Test (AUC: ', round(roc.test$auc, 3), ')'))) +
      labs(x = 'False Positive', y = 'True Positive') +
      theme_ws()
    theme_simple(legend.position = c(.2, .85))
  }
  
  print(p)
}
# perf.auc(model = cox.conv.aic, train = dt.conv.tr, test = dt.conv.test)
# perf.auc(model = lg.prog.woe.aic, train = dt.train, test = dt.test)

perf.decile <- function(actual, pred, plot = TRUE, add.legend = TRUE) {
  # check the model performance based on actual and predicted rates
  # Args:
  #    actual: a vector containing the actual status for each record
  #    pred: predicted probability for each record
  # Return:
  #    plot the figure of model performance by decile
  #    return the actual and predicted rates for each decile
  rate <- data.frame(Actual = actual , Predict = pred) %>%
    arrange(Predict) %>%
    mutate(Decile = rep(1:10, table(cut(1:nrow(.), 10)))) %>%
    group_by(Decile) %>%
    summarise(Actual.rate = mean(Actual)* 100, Predict.rate = mean(Predict) * 100,
      Freq.1 = sum(Actual), Freq.0 = n() - Freq.1,  Freq.group = n())
  
  min.xy <- min(rate[, c('Predict.rate', 'Actual.rate')])
  max.xy <- max(rate[, c('Predict.rate', 'Actual.rate')])
  
  p <- ggplot(rate, aes(x = Actual.rate, y = Predict.rate)) +
    geom_point(aes(color = as.factor(Decile)), size = 4, show.legend = add.legend) +
    geom_abline(slope = 1, linetype = 2) +
    coord_equal() +
    scale_x_continuous(limits = c(min.xy, max.xy)) +
    scale_y_continuous(limits = c(min.xy, max.xy)) +
    scale_color_brewer(name = 'Decile', palette = 'RdYlBu', direction = -1) +
    labs(x = 'Actual Rate (%)', y = 'Predicted Rate (%)') +
    guides(color = guide_legend(reverse = T, override.aes = list(size = 5))) +
    theme_bw()
  
  if(plot == TRUE) print(p)
  return(rate)
}
# perf.decile(actual = dt$Progression, pred = mod$fitted.values)

## calculate the survival probability at each time for each record one by one
survexp.obo <- function(data, ratetable, ...) {
  # calculate the survival table for each record one by one
  # Args:
  #    the same as the survexp model, except that the formula is not required
  # Return:
  #    suvival function at each time step, for each record
  
  id = split(1:nrow(data), cut(1:nrow(data), ceiling(nrow(data) / 5000)))
  
  pred.obo <- lapply(id, function(x) {
    pred <- survexp(~ ID, ratetable = ratetable,
      data = data.frame(data[x, ], ID = x), ...)
    t(pred$surv)
  })
  
  pred.all <- do.call(rbind, pred.obo)
}
# rs <- survexp.obo(data = dt.conv.test, ratetable = cox.conv.train)

## conver the regression coefficients to meaningful rates
coef2rate <- function(data, model, level.stat.output, force.change = TRUE,
  time = NULL) {
  xs <- labels(model$terms)
  
  pred.x.list <- lapply(xs, function(x) {
    
    if(force.change == TRUE) {
      groups = unique(data[, x])
      pred <- sapply(groups, function(y) {
        dt.temp <- data
        dt.temp[x] <- y
        if(class(model)[1] == 'coxph') {
          if(is.null(time)) stop('Prediction time is needed for coxph model')
          pred <- survexp(formula = ~ 1, data = dt.temp, ratetable = model,
            times = time)
          return(1 - pred$surv)
        } else {
          pred <- predict(model, newdata = dt.temp, type = 'response')
          return(mean(pred))
        }
      })
      pred.x <- data.frame(Variable = x, Group = groups, Pred.Rate.1 = pred)
    }
    
    if(force.change == FALSE) {
      if(class(model)[1] == 'coxph') {
        if(is.null(time)) stop('Prediction time is needed for coxph model')
        
        pred <- survexp(formula = as.formula(paste('~', x)), data = data,
          ratetable = model, times = time)
        
        pred.x <- data.frame(Variable = x,
          Group = gsub(paste0(x, '='), '', names(pred$surv)),
          Pred.Rate.1 = 1 - pred$surv)
        # the result is exactly the same as prediction using surexp.obo
        # then calculate the mean value for each group
        # pred.obo <- survexp.obo(data = data, ratetable = model)
        # pred.12 <- data.frame(College = data$College, Pred = pred.obo[, 12])
        # group_by(pred.12, College) %>% summarise(P = 1 - mean(Pred))
      } else {
        pred <- predict(model, newdata = data, type = 'response')
        pred.x <- data.frame(
          Variable = x, Group = data[, x], Pred.Rate.1 = pred) %>%
          group_by(Variable, Group) %>%
          summarise(Pred.Rate.1 = mean(Pred.Rate.1))
      }
    }
    return(pred.x)
  })
  
  pred.xs <- do.call(rbind, pred.x.list)
  if(all(pred.xs$Group %in% level.stat.output$WOE)) {
    pred.xs <- rename(pred.xs, WOE = Group)
  }
  
  # in case the _woe columns are used for model, remove the _woe to match
  # with the original variable names
  pred.xs$Variable <- gsub('_woe', '', pred.xs$Variable)
  
  pred.stat <- left_join(level.stat.output, pred.xs) %>%
    filter(!is.na(Pred.Rate.1)) %>%
    mutate(Variable = factor(Variable, levels = unique(Variable)),
      Group = factor(Group, levels = unique(Group)),
      Variable.IV = factor(Variable.IV, levels = unique(Variable.IV)),
      Group = factor(Group,
        levels = c(setdiff(unique(Group), 'Missing'), 'Missing'))) %>%
    data.frame
  
  return(pred.stat)
}
# conv.lvs <- level.stat(data = dt.conv, x = col.x, y = 'Conversion_Status_1yr')
# conv.pred.stat <- coef2rate(data = dt.conv, model = cox.aic,
#   level.stat.output = conv.lvs, time = 12, force.change = TRUE)

################################## plot ########################################
## ggplot themes
# a classic simple theme with x and y axis
theme_simple <- function(...) {
  theme_bw() +
    theme(axis.line.x = element_line(size = .5),
      axis.line.y = element_line(size = .5),
      strip.text = element_text(face = 'bold'),
      panel.border = element_blank(),
      strip.background = element_blank(),
      axis.title = element_text(face = 'bold'),
      axis.title.y=element_text(margin=margin(0,10,0,0)),
      legend.title = element_text(face = 'bold'),
      panel.grid.major = element_line(linetype = 'dashed'),
      panel.grid.minor = element_blank(),
      ...)
}

# a ggplot theme similar to the figures published on the Wall Street Journal
theme_ws <- function(background = 'ivory', ...) {
  theme_bw() +
    theme(axis.line.x = element_line(size = .5),
      rect = element_rect(fill = background, linetype = 0, color = NA),
      panel.border = element_blank(),
      strip.text = element_text(family = 'sans', face = 'bold.italic'),
      strip.background = element_blank(),
      axis.title = element_text(face = 'bold'),
      axis.title.y=element_text(margin=margin(0,10,0,0)),
      legend.title = element_text(face = 'bold'),
      legend.position = 'top',
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(linetype = 'dashed', color = 'grey50'),
      axis.ticks.y = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = background),
      ...)
}
# rr <- ggplot(mtcars) + geom_point(aes(x = wt, y = mpg, colour=factor(gear)))
# rr + theme_ws()
# rr + facet_wrap(~carb, ncol = 2, scale = 'free')
# rr + theme_simple() + scale_y_continuous(limits = c(0, 35))

display.col <- function() {
  # Display all possible colors in ggplot2
  # reference to http://sape.inf.usi.ch/quick-reference/ggplot2/colour
  d = grep('gray|grey.*[1-9]$|grey100', colors(), value = T, invert = T) %>%
    matrix(ncol = 9) %>%
    melt
  colnames(d) <- c('y', 'x', 'c')
  d$y = max(d$y) - d$y
  
  ggplot(data = d) +
    scale_x_continuous(name = '', breaks = NULL, expand = c(0, 0)) +
    scale_y_continuous(name = '', breaks = NULL, expand = c(0, 0)) +
    scale_fill_identity() +
    geom_rect(aes(
      xmin = x, xmax = x + 1, ymin = y, ymax = y + 1), fill = "white") +
    geom_rect(aes(
      xmin = x + 0.05, xmax = x + 0.95, ymin = y + 0.5, ymax = y + 1, fill=c)) +
    geom_text(aes(
      x = x + 0.5, y = y, label = c), nudge_y = .55, color = 'black', size=3) +
    theme_ws()
}

display.pch <- function() {
  # Display all possible pch (point shape) used for ggplot
  # reference to http://sape.inf.usi.ch/quick-reference/ggplot2/shape
  d=data.frame(p=c(0:25, 32:127))
  ggplot(data=d) +
    scale_y_continuous(name = '', breaks = NULL) +
    scale_x_continuous(name = '', breaks = NULL) +
    scale_shape_identity() +
    geom_point(aes(x=p%%16, y=7 - p%/%16, shape=p), size=5, fill="red") +
    geom_text(aes(x=p%%16, y=7 - p%/%16+0.25, label=p), size=3) +
    theme_ws()
}

## beatufied correlation plot
corrplot.beautify <- function(cor.mat) {
  # The layout and fonts of the default corrplot output doesn't look good
  # this function is used to beautify the corrplot
  # Arg:
  #    cor.mat: correlation matrix generated by cor() function
  par(cex = .8)
  corrplot(round(cor.mat, 2), type = 'lower', tl.srt = 15, addCoef.col = "black",
    cl.cex = 1.5, tl.cex = 1.2, tl.col = 'black', mar = c(0, 0, 0, 0),
    col=colorRampPalette(c("blue","white","red"))(200))
  par(cex = 1)
}
```


## Packages Setup
<ul class = "incremental">
**Install Dependent Packages**
```{r, eval=FALSE}
# If the default mirror is blocked, choose another mirror to install R packages, 
# chooseCRANmirror()
sapply(c('dplyr', 'car', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart', 
  'scales', 'shiny', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS', 'pROC', 
  'manipulate'), 
  install.packages)
```

**Load Dependent Packages**
```{r}
# Load pacakges
# sapply(c('dplyr', 'car', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart', 
#   'scales', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS', 'pROC', 
#   'manipulate'), 
#   require, character.only = TRUE)

# shinyapp can not recognizes the libraries needed if they are load all together  
# with the sapply function above. The library must be load one by one for shinyapp!
library(dplyr)
library(car)
library(caret)
library(e1071)
library(knitr)
library(reshape2)
library(corrplot)
library(rpart)
library(scales)
library(survival)
library(gridExtra)
library(devtools)
library(pec)
library(MASS)
library(pROC)
library(manipulate)
```

**Install `streamlineR` from Github**
```{r,eval=FALSE}
# Ff the install_github does not work, you can download the package from github,
# and then install the package locally: 
# https://api.github.com/repos/JianhuaHuang/streamlineR/zipball/master 
devtools::install_github('JianhuaHuang/streamlineR')
library(streamlineR)
```
</ul>

## Data Preparation
<ul class = "incremental">
**Load Data**
```{r}
dt <- survival::pbc %>%
  transmute(age = round(age), gender = sex, platelet, stage = as.character(stage), 
    time, status = as.numeric(status %in% c(1, 2))) %>%
  filter(!is.na(stage))

set.seed(1111)  # reproducible
dt <- dt[sample(nrow(dt), 10000, replace = T), ]
row.names(dt) <- NULL
dim(dt)
str(dt)
head(dt)
```

**Split Data into Training and Test Datasets**
```{r}
set.seed(1111)
ind.train <- createDataPartition(dt$status, p = .7, list = FALSE)
dt.train <- dt[ind.train, ]
dt.test <- dt[-ind.train, ]
row.names(dt.train) <- NULL
row.names(dt.test) <- NULL
dim(dt.train)
dim(dt.test)

# preserve the original values
dt.train.bk <- dt.train
dt.test.bk <- dt.test
```
</ul>


## Bin Training Data Based on Regression Coefficients: `bin.knn`
<ul class = "incremental">
```{r}
bin.knn(status ~ platelet, data = dt.train, n.group = 5, min.bucket = 0.05)
```


```{r,eval=FALSE}
manipulate(bin.knn(status ~ platelet, data = dt.train, n.group, min.bucket),
  n.group = slider(1, 10, step = 1, initial = 5, label = 'Number of Groups'),
  min.bucket = slider(0.01, .2, step = 0.01, initial = 0.05,
    label = 'Minimum Bucket'))
```


```{r,echo=FALSE}
inputPanel(
  sliderInput("n_group", label = 'Number of Groups',
              min = 1, max = 9, value = 5, step = 1, ticks = FALSE),
  sliderInput("min_pop", label = 'Minimum Bucket',
              min = 0.01, max = .2, value = .05, step = .01, ticks = FALSE)
)

renderPlot({
  bin.knn(formula = status ~ platelet, data = dt.train.bk, 
    n.group = input$n_group, min.bucket = input$min_pop)
})
```
*Sometimes the web server is not stable. Please reload/refresh the page, if you can not see a figure above*
</ul>



## Bin Training Data Based on rpart: `bin.rpart` 
<ul class = "incremental">
**Decision Tree Algorithm (Recursive Partitioning)**: `rpart`
```{r}
rpart(formula = status ~ age, data = dt.train, 
  control = rpart.control(minbucket = .01 * nrow(dt.train)))
```

**Binning for Logistic Model**
```{r}
lg.bin.age <- bin.rpart(formula = status ~ age, data = dt.train, 
  rcontrol = rpart.control(minbucket = .01 * nrow(dt.train)))

str(lg.bin.age)

lg.bin.platelet <- bin.rpart(formula = status ~ platelet, data = dt.train, 
  rcontrol = rpart.control(minbucket = .01 * nrow(dt.train)))
```


**Binning for Survival Model**
```{r}
surv.bin.age <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .01 * nrow(dt.train)))  ## cp = 0.01
```


**Search for Appropriate Number of Cut Points**
```{r,echo=FALSE,fig.width=6}
inputPanel(
  sliderInput("cp", label = 'cp (complexity parameter)', min = 0.001, max = .03,
    value = .01, step = .001, ticks = FALSE)
)

renderPrint({
  rs <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train.bk,
    rcontrol = rpart.control(cp  = input$cp, minbucket = .01 * nrow(dt.train.bk))) 
  # print(rs$cut.points)
})
```
*Sometimes the web server is not stable. Please reload/refresh the page, if you can not see cut points above*

```{r}
surv.bin.age2 <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .01 * nrow(dt.train)), n.group = 3:7)
```


**Replace Numerical Varialbes with Bins**
```{r}
# We don't need the time column anmore, delete it in both dt.train and dt.test
dt.train <- dplyr::select(dt.train, -time)  
dt.test <- dplyr::select(dt.test, -time)
head(dt.train)
dt.train$age <- lg.bin.age$bins
dt.train$platelet <- lg.bin.platelet$bins
head(dt.train)
```
</ul>


## Level Statistics (Frequence, Rate, WOE, and IV): `level.stat`
<ul class = "incremental">
```{r}
col.x <- c('age', 'gender', 'platelet', 'stage')
stat.train <- level.stat(dt.train, x = col.x, y = 'status')
head(stat.train)
```
</ul>

## Visualize Level Statistics: `ggstat`
<ul class = "incremental">
**Plot with Default Arguments**
```{r}
ggstat(data = stat.train, var = 'Variable.IV', x = 'Group', y = 'Rate.1',
  y.label = 'Perc.1', y.min.0 = FALSE, y.title = NULL, bar.width = 'Rate.group',
  bar.width.label = 'Perc.group', n.col = NULL)
```

**Constant Bar Width**
```{r}
ggstat(stat.train, bar.width = NULL)
```

**Plot WOE**
```{r,fig.height=3}
ggstat(stat.train, y = 'WOE', y.label = 'WOE.round', bar.width = NULL, 
  bar.width.label = NULL, n.col = 4)
```
</ul>

## Replace Bins with WOE: `replace.woe`
<ul class = "incremental">
```{r}
replace.woe(data = dt.train, level.stat.output = stat.train, replace = FALSE) %>%
  head

dt.train <- replace.woe(data = dt.train, level.stat.output = stat.train, 
  replace = TRUE)
head(dt.train)
```
</ul>

## Correlation between Independent Variables: `corrplot.beautify`
<ul class = "incremental">
```{r}
cor.mat <- cor(dt.train[, col.x])
corrplot(cor.mat)
corrplot.beautify(cor.mat)
```
</ul>

## Logistic Model
<ul class = "incremental">
**Full Model**
```{r}
lg <- glm(status ~ ., dt.train, family=binomial(link='logit'))
summary(lg)
```

**Stepwise Variable Selection**
```{r}
lg.aic <- stepAIC(lg, k =  qchisq(0.05, 1, lower.tail=F))   # p to enter: 0.05
summary(lg.aic)

data.frame(vif(lg.aic))
```
</ul>

## Prepare Test Data: `bin.custom & replace.woe`
<ul class = "incremental">
**Bin Test Data**: `bin.custom`
```{r}
head(dt.test)
dt.test$age <-  bin.custom(dt.test$age, cut.p = lg.bin.age$cut.points)
dt.test$platelet <- bin.custom(dt.test$platelet, cut.p = lg.bin.platelet$cut.points)
head(dt.test)
```


**Replace Binned Test Data with WOE**: `replace.woe`
```{r}
dt.test <- replace.woe(dt.test, level.stat.output = stat.train, replace = TRUE)
head(dt.test)
```
</ul>

## Check Model Performance: `perf.auc & perf.decile`
<ul class = "incremental">
**Check Performance Based on AUC**: `perf.auc`
```{r}
perf.auc(model = lg.aic, dt.train, dt.test)
```


**Check Performance Based on Decile Rate**: `perf.decile`
```{r}
pred.test <- predict(lg.aic, newdata = dt.test, type = 'response')
perf.decile(actual = dt.test$status, pred = pred.test, add.legend = TRUE)
```
</ul>


## Convert Coefficients to Rate: `coef2rate`
<ul class = "incremental">
```{r}
summary(lg.aic)
pred.stat <- coef2rate(data = dt.test, model = lg.aic, 
  level.stat.output = stat.train, force.change = TRUE)
head(pred.stat)
# Compare The Actual & Prediction Rates
pred.stat[,c('Rate.1', 'Pred.Rate.1')]

ggstat(pred.stat, y = 'Pred.Rate.1')
```
</ul>

## Reference:
* streamlineR package information: https://github.com/JianhuaHuang/streamlineR 
* Submit bug: http://github.com/JianhuaHuang/streamlineR/issues


