---
title: 'Streamline Routine Modeling Work in R: streamlineR'
author: "Jianhua Huang"
date: "July 27, 2016"
output: 
  slidy_presentation:
    fig_width: 7
    fig_height: 5
    duration: 60
    incremental: true
runtime: shiny
---

```{r knitr.setup, include=FALSE}
rm(list = ls())
# dir <- 'F:/Projects/Rpackage/streamlineR'
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


## Packages Setup
<ul class = "incremental">
**Install Dependent Packages**
```{r, eval=FALSE}
# If the default mirror is blocked, choose another mirror to install R packages, 
# chooseCRANmirror()
sapply(c('dplyr', 'car', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart', 
  'scales', 'shiny', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS', 'pROC', 
  'manipulate'), 
  install.packages)
```

**Load Dependent Packages**
```{r}
# Load pacakges
sapply(c('dplyr', 'car', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart', 
  'scales', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS', 'pROC', 
  'manipulate'), 
  require, character.only = TRUE)
```

**Install `streamlineR` from Github**
```{r}
# Ff the install_github does not work, you can download the package from github,
# and then install the package locally: 
# https://api.github.com/repos/JianhuaHuang/streamlineR/zipball/master 
devtools::install_github('JianhuaHuang/streamlineR')
library(streamlineR)
```
</ul>

## Data Preparation
<ul class = "incremental">
**Load Data**
```{r}
dt <- survival::pbc %>%
  transmute(age = round(age), gender = sex, platelet, stage = as.character(stage), 
    time, status = as.numeric(status %in% c(1, 2))) %>%
  filter(!is.na(stage))

set.seed(1111)  # reproducible
dt <- dt[sample(nrow(dt), 10000, replace = T), ]
row.names(dt) <- NULL
dim(dt)
str(dt)
head(dt)
```

**Split Data into Training and Test Datasets**
```{r}
set.seed(1111)
ind.train <- createDataPartition(dt$status, p = .7, list = FALSE)
dt.train <- dt[ind.train, ]
dt.test <- dt[-ind.train, ]
row.names(dt.train) <- NULL
row.names(dt.test) <- NULL
dim(dt.train)
dim(dt.test)

# preserve the original values
dt.train.bk <- dt.train
dt.test.bk <- dt.test
```
</ul>


## Bin Training Data Based on Regression Coefficients: `bin.knn`
<ul class = "incremental">
```{r}
bin.knn(status ~ platelet, data = dt.train, n.group = 5, min.bucket = 0.05)
```


```{r,eval=FALSE}
manipulate(bin.knn(status ~ platelet, data = dt.train, n.group, min.bucket),
  n.group = slider(1, 10, step = 1, initial = 5, label = 'Number of Groups'),
  min.bucket = slider(0.01, .2, step = 0.01, initial = 0.05,
    label = 'Minimum Population'))
```


```{r,echo=FALSE}
inputPanel(
  sliderInput("n_group", label = 'Number of Groups',
              min = 1, max = 9, value = 5, step = 1, ticks = FALSE),
  sliderInput("min_pop", label = 'Minimum Population',
              min = 0.01, max = .2, value = .05, step = .01, ticks = FALSE)
)

renderPlot({
  bin.knn(formula = status ~ platelet, data = dt.train.bk, 
    n.group = input$n_group, min.bucket = input$min_pop)
})
```
</ul>

## Bin Training Data Based on rpart: `bin.rpart` 
<ul class = "incremental">
**Decision Tree Algorithm (Recursive Partitioning)**: `rpart`
```{r}
rpart(formula = status ~ age, data = dt.train, 
  control = rpart.control(minbucket = .01 * nrow(dt.train)))
```

**Binning for Logistic Model**
```{r}
lg.bin.age <- bin.rpart(formula = status ~ age, data = dt.train, 
  rcontrol = rpart.control(minbucket = .01 * nrow(dt.train)))

str(lg.bin.age)

lg.bin.platelet <- bin.rpart(formula = status ~ platelet, data = dt.train, 
  rcontrol = rpart.control(minbucket = .01 * nrow(dt.train)))
```


**Binning for Survival Model**
```{r}
surv.bin.age <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)))  ## cp = 0.01
```


**Search for Appropriate Number of Cut Points**
```{r,echo=FALSE,fig.width=6}
inputPanel(
  sliderInput("cp", label = 'cp', min = 0.001, max = .03, value = .01, 
    step = .001, ticks = FALSE)
)

renderPrint({
  rs <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train.bk,
    rcontrol = rpart.control(cp  = input$cp, minbucket = .05 * nrow(dt.train.bk))) 
  # print(rs$cut.points)
})
```


```{r}
surv.bin.age2 <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)), n.group = 3:7)
```


**Replace Numerical Varialbes with Bins**
```{r}
# We don't need the time column anmore, delete it in both dt.train and dt.test
dt.train <- dplyr::select(dt.train, -time)  
dt.test <- dplyr::select(dt.test, -time)
head(dt.train)
dt.train$age <- lg.bin.age$bins
dt.train$platelet <- lg.bin.platelet$bins
head(dt.train)
```
</ul>


## Level Statistics (Frequence, Rate, WOE, and IV): `level.stat`
<ul class = "incremental">
```{r}
col.x <- c('age', 'gender', 'platelet', 'stage')
stat.train <- level.stat(dt.train, x = col.x, y = 'status')
head(stat.train)
```
</ul>

## Visualize Level Statistics: `ggstat`
<ul class = "incremental">
**Plot with Default Arguments**
```{r}
ggstat(data = stat.train, var = 'Variable.IV', x = 'Group', y = 'Rate.1',
  y.label = 'Perc.1', y.min.0 = FALSE, y.title = NULL, bar.width = 'Rate.group',
  bar.width.label = 'Perc.group', n.col = NULL)
```

**Constant Bar Width**
```{r}
ggstat(stat.train, bar.width = NULL)
```

**Plot WOE**
```{r,fig.height=3}
ggstat(stat.train, y = 'WOE', y.label = 'WOE.round', bar.width = NULL, 
  bar.width.label = NULL, n.col = 4)
```
</ul>

## Replace Bins with WOE: `replace.woe`
<ul class = "incremental">
```{r}
replace.woe(data = dt.train, level.stat.output = stat.train, replace = FALSE) %>%
  head

dt.train <- replace.woe(data = dt.train, level.stat.output = stat.train, 
  replace = TRUE)
head(dt.train)
```
</ul>

## Correlation between Independent Variables: `corrplot.beautify`
<ul class = "incremental">
```{r}
cor.mat <- cor(dt.train[, col.x])
corrplot(cor.mat)
corrplot.beautify(cor.mat)
```
</ul>

## Logistic Model
<ul class = "incremental">
**Full Model**
```{r}
lg <- glm(status ~ ., dt.train, family=binomial(link='logit'))
summary(lg)
```

**Stepwise Variable Selection**
```{r}
lg.aic <- stepAIC(lg, k =  qchisq(0.05, 1, lower.tail=F))   # p to enter: 0.05
summary(lg.aic)

data.frame(vif(lg.aic))
```
</ul>

## Prepare Test Data: `bin.custom & replace.woe`
<ul class = "incremental">
**Bin Test Data**: `bin.custom`
```{r}
head(dt.test)
dt.test$age <-  bin.custom(dt.test$age, cut.p = lg.bin.age$cut.points)
dt.test$platelet <- bin.custom(dt.test$platelet, cut.p = lg.bin.platelet$cut.points)
head(dt.test)
```


**Replace Binned Test Data with WOE**: `replace.woe`
```{r}
dt.test <- replace.woe(dt.test, level.stat.output = stat.train, replace = TRUE)
head(dt.test)
```
</ul>

## Check Model Performance: `perf.auc & perf.decile`
<ul class = "incremental">
**Check Performance Based on AUC**: `perf.auc`
```{r}
perf.auc(model = lg.aic, dt.train, dt.test)
```


**Check Performance Based on Decile Rate**: `perf.decile`
```{r}
pred.test <- predict(lg.aic, newdata = dt.test, type = 'response')
perf.decile(actual = dt.test$status, pred = pred.test, add.legend = TRUE)
```
</ul>


## Convert Coefficients to Rate: `coef2rate`
<ul class = "incremental">
```{r}
summary(lg.aic)
pred.stat <- coef2rate(data = dt.test, model = lg.aic, 
  level.stat.output = stat.train, force.change = TRUE)
head(pred.stat)
# Compare The Actual & Prediction Rates
pred.stat[,c('Rate.1', 'Pred.Rate.1')]

ggstat(pred.stat, y = 'Pred.Rate.1')
```
</ul>

## Reference:
* streamlineR package information: https://github.com/JianhuaHuang/streamlineR 
* Submit bug: http://github.com/JianhuaHuang/streamlineR/issues


