---
title: 'Streamline Rountine Modeling Work in R: streamlineR'
author: "Jianhua Huang"
date: "July 27, 2016"
output: 
  slidy_presentation:
    fig_width: 7
    fig_height: 4
    duration: 60
runtime: shiny
---

```{r setup, include=FALSE}
rm(list = ls())
# If you don't have these packages, install them. 
# once the packages are installed, you don'Bt need to run the code again
# sapply(c('dplyr', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart',
#   'scales', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS'), install.packages)

# Load these pacakges
sapply(c('dplyr', 'caret', 'e1071', 'knitr', 'reshape2', 'corrplot','rpart', 
  'scales', 'survival', 'gridExtra', 'devtools', 'pec', 'MASS', 'pROC', 
  'manipulate'), 
  require, character.only = TRUE)

# dir <- 'F:/Projects/Rpackage/streamlineR'
# source('F:/Projects/Rpackage/streamlineR/R/myFun.R')
dir <- 'C:/Users/Jianhua/Dropbox/work_doc/Rpackage/streamlineR'
# source('C:/Users/Jianhua/Dropbox/work_doc/Rpackage/streamlineR/R/myFun.R')
opts_chunk$set(echo = TRUE, root.dir = dir, warning=FALSE, message=FALSE)
```


## Data Preparation
```{r}
# chooseCRANmirror()  # choose another mirror to download R packages, if blocked
install_github('JianhuaHuang/streamlineR')
library(streamlineR)

dt <- survival::pbc %>%
  transmute(age = round(age), gender = sex, platelet, stage = as.character(stage), 
    time, status = as.numeric(status %in% c(1, 2))) %>%
  filter(!is.na(stage))

set.seed(1111)
dt <- dt[sample(nrow(dt), 10000, replace = T), ]
dim(dt)
str(dt)
head(dt)
```


## Split Data into Training and Test datasets
```{r}
# set.seed(123456)
set.seed(1111)
ind.train <- createDataPartition(dt$status, p = .7, list = FALSE)
dt.train <- dt[ind.train, ]
dt.test <- dt[-ind.train, ]
row.names(dt.train) <- 1:nrow(dt.train)
row.names(dt.test) <- 1:nrow(dt.test)
dim(dt.train)
dim(dt.test)
rm(dt)
```


## Binning Based on Regression Coefficients: `bin.knn`
```{r,eval=FALSE}
manipulate(bin.knn(status ~ platelet, data = dt.train, n.group, min.pop),
  n.group = slider(1, 10, step = 1, initial = 5, label = 'Number of Groups'),
  min.pop = slider(0.01, .2, step = 0.01, initial = 0.05,
    label = 'Minimum Population'))
```

```{r,echo=FALSE}
bin.knn(formula = status ~ platelet, data = dt.train, n.group = 5, min.pop = 0.05)
```


## Binning Based on rpart: `bin.rpart`
```{r}
rpart(formula = status ~ age, data = dt.train, 
  control = rpart.control(minbucket = .05 * nrow(dt.train)))
```

**Binning for Logistic Model**
```{r}
lg.bin.age <- bin.rpart(formula = status ~ age, data = dt.train, 
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)))

str(lg.bin.age)

lg.bin.platelet <- bin.rpart(formula = status ~ platelet, data = dt.train, 
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)))
```

**Binning for Survival Model**
```{r}
surv.bin.age <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)))  ## cp = 0.01

surv.bin.age <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(cp  = .001, minbucket = .05 * nrow(dt.train)))

surv.bin.age2 <- bin.rpart(formula = Surv(time, status) ~ age, data = dt.train,
  rcontrol = rpart.control(minbucket = .05 * nrow(dt.train)), n.group = 3:7)
```


```{r}
inputPanel(
  sliderInput("cp", label = 'cp',
              min = 0.001, max = .1, value = .01, step = .001)
)

renderPrint({
  rs <- bin.rpart(formula = status ~ platelet, data = dt.train,
    rcontrol = rpart.control(cp  = input$cp, minbucket = 100)) 
  print(rs$cut.points)
})
```


**Replace numerical Varialbes with Bins**
```{r}
dt.train <- dplyr::select(dt.train, -time)
dt.test <- dplyr::select(dt.test, -time)
head(dt.train)
dt.train$age <- lg.bin.age$bins
dt.train$platelet <- lg.bin.platelet$bins
head(dt.train)
```


## Level Statistics (Frequence, Rate, WOE, and IV): `level.stat`
```{r}
col.x <- c('age', 'gender', 'platelet', 'stage')
stat.train <- level.stat(dt.train, x = col.x, y = 'status')
head(stat.train)
```


## Visualizing Level Statistics: `ggstat`
```{r}
ggstat(data = stat.train, var = 'Variable.IV', x = 'Group', y = 'Rate.1',
  y.label = 'Perc.1', y.min.0 = FALSE, y.title = NULL, bar.width = 'Rate.group',
  bar.width.label = 'Perc.group', n.col = NULL)

ggstat(stat.train)
```


**Constant Bar Width**
```{r}
ggstat(stat.train, bar.width = NULL)
```


**Plot WOE**
```{r,fig.height=3}
stat.train$WOE.round <- round(stat.train$WOE, 2)
ggstat(stat.train, y = 'WOE', y.label = 'WOE.round', bar.width = NULL, 
  bar.width.label = NULL, n.col = 4)
```


## Replace Bins with WOE: `replace.woe`
```{r}
replace.woe(data = dt.train, level.stat.output = stat.train) %>% head

dt.train.org <- dt.train
dt.train <- replace.woe(data = dt.train, level.stat.output = stat.train, 
  replace = TRUE)
head(dt.train)
```


## Correlation between Independent Variables: `corrplot.beautify`
```{r}
cor.mat <- cor(dt.train[, col.x])
corrplot.beautify(cor.mat)
```


## Logistic Model
```{r}
lg <- glm(status ~ ., dt.train, family=binomial(link='logit'))
summary(lg)

lg.aic <- stepAIC(lg, k =  qchisq(0.05, 1, lower.tail=F))   # p to enter: 0.05
summary(lg.aic)
```


## Preparing Test Data: `bin.custom & replace.woe`
**Bin Test Data**: `bin.custom`
```{r}
head(dt.test)
dt.test$age <-  bin.custom(dt.test$age, cut.p = lg.bin.age$cut.points)
dt.test$platelet <- bin.custom(dt.test$platelet, cut.p = lg.bin.platelet$cut.points)
head(dt.test)
```


**Replace Binned Test Data with WOE**: `replace.woe`
```{r}
dt.test <- replace.woe(dt.test, level.stat.output = stat.train, replace = TRUE)
head(dt.test)
```


## Model Performance: `perf.auc & perf.decile`
**Check Performance Based on AUC**: `perf.auc`
```{r}
perf.auc(model = lg.aic, dt.train, dt.test)
```


**Check Performance Based on Decile Rate**: `perf.decile`
```{r}
pred.test <- predict(lg.aic, newdata = dt.test, type = 'response')
perf.decile(actual = dt.test$status, pred = pred.test, add.legend = TRUE)
```


## Convert Coefficients to Rate: `coef2rate`
```{r}
pred.stat <- coef2rate(data = dt.test, model = lg.aic, 
  level.stat.output = stat.train, force.change = TRUE)
head(pred.stat)
pred.stat[,c('Rate.1', 'Pred.Rate.1')]

ggstat(pred.stat, y = 'Pred.Rate.1')

inputPanel(
  sliderInput("gg_n_group", label = "Number of Groups",
              min = 1, max = 4, value = 2, step = 1)
)

renderPlot({
  ggstat(stat.train, n.col = input$gg_n_group)
})

```

